Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2014-10-06 article.Rmd
More conclusion

Our algorithm for predicting gender from first names, when combined with historical datasets in the Gender package, provides a powerful new method for researchers interested in the study of gender. The digital turn has made available very large datasets, yet in the midst of abundance many humanists continue to deal with the problem of incompleteness and scarcity within their sources. Our method uses *abundant* existing datasets to infer additional information from *scarce* datasets. The temporal dimension of our work matches historical questions to a historical method, providing much improved results over simple, anachronistic, or ahistorical lookup methods. This method is also extensible to different times and places, provided a researcher has a similar reference dataset of names and genders from the period or area under study.

The analytical category of gender has transformed research in the humanities for the past half-century. More recently, the digitization movement has opened up exciting opportunities to interrogate gender in new ways. Projects in the digital humanities have ranged from recovering the voices of early-modern women writers to tracing the linguistic styles of male and female authors [Women Writers Project n.d., Argamon et. al. 2009]. The Gender package is a new tool to add to this approach. It reverses the traditional disciplinary direction bewteen computer science and the humanities. Rather than adapting existing computational methodologies, the Gender package uses a humanistic starting point---the historical relationship between gender and naming practices---to contribute a new method to computer science.

>>>>

Our algorithm for predicting gender from first names, when combined with historical datasets in the Gender package, provides a powerful new method for researchers interested in the study of gender. The digital turn has made available very large datasets, yet in the midst of abundance many humanists continue to deal with the problem of incompleteness and scarcity within their sources. Our method uses *abundant* existing datasets to infer additional information from *scarce* datasets. The temporal dimension of our work matches historical questions to a historical method, providing much improved results over simple, anachronistic, or ahistorical lookup methods. This method is also extensible to different times and places, provided a researcher has a similar reference dataset of names and genders from the period or area under study. The Gender package reverses the traditional disciplinary direction bewteen computer science and the humanities. Rather than adapting existing computational methodologies, it uses a humanistic starting point---the historical relationship between gender and naming practices---to contribute a new method to computer science.
The analytical category of gender has transformed research in the humanities for the past half-century. More recently, the digitization movement has opened up exciting opportunities to interrogate gender in new ways. Projects in the digital humanities have ranged from recovering the voices of early-modern women writers to tracing the linguistic styles of male and female authors [Women Writers Project n.d., Argamon et. al. 2009]. The proliferation of machine-readable datasets and digital texts brings a vast trove of material with which to study gender. But unlocking its full range of meaning necessitates not just accessing the data that *is* there, but inferring information that *isn't* there. The Gender package is a powerful tool to do just that.
